# Introduction  {#intro_eval}

```{r setup_eval, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Generalities

The objective of the surveillance program is to demonstrate the freedom of the Swiss cattle population from Infectious bovine rhinotracheitis (IBR) and Enzootic bovine leukosis (EBL) in accordance with the requirements of the bilateral treaties with the EU.

## Methods

The evaluation of the surveillance system is used to assess the probability of freedom from IBR and EBL in Swiss livestock herds given a certain threshold for inter-herd prevalence $P_{betH}$ and intra-herd prevalence $P_{inH}$.

Dairy and non-dairy (furtherby named beef) herds are considered as separate populations. Each of these two subpopulations includes two types of samples (random samples and sentinel samples, where farms belonging to the latter are called sentinel farms). The method used is based on the one described by [@RN2]. 

Thus, monitoring is performed in a total of four compartments (dairy herds with random sampling, risk based sampling of dairy sentinel herds, beef herds with random sampling and risk-based sampling of beef sentinel herds). 
Dairy farms are sampled by tank milk sampling while beef farms are sampled at the slaughterhouse while for beef farms blood samples from single animals are collected. No risk factors are considered in the random component. The sentinel component includes sentinel herds selected based on their relative risk. The risk factors considered are: **animal contacts with foreign livestock operations**, **above average animal traffic**, **high density of livestock operations in the neighbourhood**, **border livestock operations** and **cattle imports**.

## Programming tools

The planning tool for the surveillance programme for IBR and EBL was coded with the language `R version 4.2.0 "Vigorous Calisthenics"` [@R-base] in the integrated development environment `RStudio version 2022.02.2 Build 485 "Prairie Trillium"`. The packages used are `tidyverse version 1.3.2` [@R-tidyverse; @tidyverse2019], which itself loads several packages which we used (`tibble version 3.1.8` [@R-tibble], `dplyr version 1.0.9` [@R-dplyr], `tidyr version 1.2.0` [@R-tidyr], `stringr version 1.4.1` [@R-stringr], `readr version 2.1.2` [@R-readr] and `purrr version 0.3.4` [@R-purrr]). Additionally, other packages from the *tidyverse* but not loaded by default by the package of the same name were used: `readxl version 1.4.1` [@R-readxl], `dtplyr version 1.2.2` [@R-dtplyr] and `lubridate version 1.8.0` [@R-lubridate; @lubridate2011]. The packages `freedom version 1.0.1` [@R-freedom] and `openxlsx version 4.2.5` [@R-openxlsx] were also used.

This documentation will not dwell long on information regarding the packages, and will only provide two things. Firstly, the code that was used to load and, if necessary, first install the packages, can be seen below. Secondly, this documentation will provide some information on the `freedom` package and explain why some adaptations were made for two functions.

### Loading of the packages {#packages_eval}

This code was used to load packages. If packages were not installed, they were first installed and then loaded. The results, messages and warnings called by the code are also provided.

```{python load_packages_eval, eval=FALSE}
packages <- c("tidyverse", "readxl", "dtplyr", "lubridate", "freedom", "openxlsx")

for (i in packages) {
  if(require(i, character.only = TRUE)) {
    print(paste(i, "is loaded correctly"))
  } else {
    print(paste("trying to install", i))
    install.packages(i, dependencies = TRUE, type = "win.binary")
    if(require(i, character.only = TRUE)) {
      print(paste(i, "installed and loaded"))
    } else {
      stop(paste("could not install", i))
    }
  }
}

remove(packages, i)
```

```{r load_other_packages_eval, include=FALSE}
# The packages "knitr", "bookdown", "kableExtra" and "xaringanExtra" were only used to create the HTML documentation, and have nothing with the evaluation code to do. Therefore, they will be loaded separately and their code will be hided.

library(tidyverse)
library(readxl)
library(dtplyr)
library(lubridate)
library(freedom)
library(openxlsx)
library(knitr)
library(bookdown)
library(kableExtra)
library(xaringanExtra)
```

### Modification of functions of the `freedom` package {#ModificationFn}

The `freedom` package implements functions used to calculate probability of disease freedom. Even if the different mathematical formula were already described before, the package builds on the publication of [@RN2], which provided a framework to aggregate data from different sources and different surveillance components together to calculate the surveillance system sensitivity $SSe$ and the probability of freedom $PostPFree$. This method is known as 'scenario-tree modelling to provide disease freedom'.

The functions `hse` calculates the herd sensitivity $HSe$. It is based on the following formula: $HSe = 1 - (1 - P_{inH}) ^ {N_{tested}}$ with $P_{inH}$ the prevalence of positive animals within the herd and $N_{tested}$ the number of animals tested within the herd. $HSe$, the herd sensitivity, represents the sensitivity of the surveillance, i.e. the probability to find at least one positive animal when testing $N_{tested}$ animals in the herd if the herd was infected with a prevalence $P_{inH}$.

However, this assumes that there is sampling with replacement. This assumption can be made when the herd is big enough and only a small proportion of animals are tested. However, if this is not the case, another formula needs to be used [@RN3]. This formula uses an approximation to the hypergeometric distribution to calculate the $HSe$.

The function `hse` uses this formula when a threshold representing the $\frac{N_{tested}}{N_{animals}}$, $N_{animals}$ representing the total number of animals in the herd, is exceeded. The default for the threshold is $0.1$. In this case, the function `hse` calls the function `hse_finite`.

The hypergeometric distribution cannot use the prevalence $P_{inH}$, but needs an integer $N_{pos}$ representing the number of positive animals in the herd. The integer is needed as well for the mathematical formula than for the biological plausibility. Indeed, an animal is either positive or negative, but cannot be part positive and part negative. The approximation of the hypergeometric distribution can accept positive, non-integer values.

$N_{pos} = P_{inH} \cdot N_{tested}$. However, with this calculation, $N_{pos}$ can also have non-integer values. [@RN3] proposed to round $N_{pos}$ up, arguing that a part-positive animal should be biologically be considered as wholly positive. The authors of the `freedom` package decided not to round $P_{inH}$ and to keep non-integer values, as rounding up could possibly lead to overestimation of $HSe$.

As an example, when comparing both approaches with a herd of 11 animals, a prevalence of 10 percent and testing 3 animals ($N_{animals} = 11$, $P_{inH} = 0.1$, $N_{tested} = 3$), the approach of [@RN3] provides $HSe = 0.451$ while the approach of the authors of the package `freedom` provides $HSe = 0.281$.

We decided to use generally the approach of the authors of the package `freedom`, as it was more conservative with the results of $HSe$. However, we decided to make a small modification to the `hse_finite` function when the expected number of positive animals would lie below 1 ($N_{pos} < 1$). Indeed, if a herd is infected, then there must have at least one positive animal within the herd. Therefore, we modified `hse_finite` in order to round $N_{pos}$ to one when below one, and to keep its non-integer value if it lies above one.

The new function `hse_finite_modified` was created by copying the original code of the `hse_finite` function of the `freedom` package and modifying it as following:

```{r hse_finite_modified_eval}
hse_finite_modified <- function(id,
                       n_tested,
                       N,
                       test_Se,
                       dp) {
  
  if (length(n_tested) != length(N)) {
    stop(paste("The length of the n_tested vector must be equal to the N",
               "vector. ie. you must describe both the number of animals",
               "tested in each group as well as how many animals are in",
               "each group.", sep = "\n"))
  }
  
  if (any(n_tested > N)) {
    stop("One of the URG has more subunits tested than in the population")
  }
  
  if (!(length(dp) == 1 || length(n_tested) == length(dp))) {
    stop(paste("The length of the n_tested vector must be equal to",
               "the dp vector. ie. you must describe both the number of",
               "animals tested in each group as well as the dp in",
               "each group.", sep = "\n"))
  }
  
  if (!(length(test_Se) == 1 || length(test_Se) == length(n_tested))) {
    stop("The length of test_Se must be either 1 or the length of n_tested")
  }
  
  if (!(length(id) == length(n_tested))) {
    stop(paste("Argument id (grouping variable) should be",
               "the same length as n_tested"))
  }
  
  A <- 1 - (n_tested * test_Se / N)
  
  B <- pmax(dp * N, 1) # Change to the original function to round to 1 if below 1
  
  df <- as.data.frame(1 - tapply(A ^ B, INDEX = id, FUN = "prod"))
  
  names(df) <- c("HSe")
  
  df$id <- rownames(df)
  
  df[, c("id", "HSe")]
}
```

In consequence, the function `hse` was also updated to `hse_modified` to call `hse_finite_modified` instead of `hse_finite` when the threshold was exceeded.

```{r hse_modified_eval}
hse_modified <- function(id,
                n_tested,
                N,
                test_Se,
                dp,
                threshold = 0.1,
                force = FALSE) {
  
  ## Ratio of animals tested in the herds
  ratio <- n_tested / N
  
  ## Check if this is more than expected
  if (any(ratio > 1) & !force) {
    problem <- id[ratio > 1]
    stop(paste("Greater than 100% of animals cannot be tested.",
               "This occurs in the following ids:",
               paste(problem, collapse = ", "),
               "To ignore this an default to infinite population",
               "for these herds, set force = TRUE", sep = "\n"))
  }
  
  ## Use the finite calculation for those with more than the threshold
  finite <- NULL
  index_finite <- (ratio > threshold) & (ratio <= 1)
  test_Se_finite <- test_Se
  if (length(test_Se) > 1) {
    test_Se_finite <- test_Se[index_finite]
  }
  dp_finite <- dp
  if (length(dp) > 1) {
    dp_finite <- dp[index_finite]
  }
  if (any(index_finite)) {
    finite <- hse_finite_modified(id[index_finite], # Change to the original function to use 'hse_finite_modified' and not 'hse_finite'
                                  n_tested[index_finite],
                                  N[index_finite],
                                  test_Se_finite,
                                  dp_finite)
    finite$method <- "finite"
  }
  if (all(index_finite)) {
    return(finite)
  }
  
  ## Otherwise use the infinite
  index_infinite <- !index_finite
  test_Se_infinite <- test_Se
  if (length(test_Se) > 1) {
    test_Se_infinite <- test_Se[index_infinite]
  }
  dp_infinite <- dp
  if (length(dp) > 1) {
    dp_infinite <- dp[index_infinite]
  }
  
  infinite <- freedom::hse_infinite(id[index_infinite],
                           n_tested[index_infinite],
                           test_Se_infinite,
                           dp_infinite)
  infinite$method <- "infinite"
  
  ## return the complete dataset
  rbind(finite,
        infinite)
}
```

## Outline of the steps to evaluate the surveillance programme

To evaluate the surveillance programme, several information are needed. First, a list of all the samples taken and tested throughout the surveillance programme must be loaded. Also, different parameters must be defined, such as the target posterior probability of freedom ($PostPFree$) or the number of tested animals per herd ($N_{tested}$). Finally, a list displaying the adjusted relative risk ($ARF$) of each of the sentinel farms must also be retrieved.

First, these data will be loaded (see [chapter 7](#load_eval)).

Then, the calculations of the component sensitivity $CSe$ for each of the components of the tree will be computed (see [section 8.1.1](#calc_cse_eval)).

Then, the sensitivity of the surveillance system $SSe$ will be calculated (see [section 8.1.2](#calc_sse_eval)).

This will allow us to calculate the posterior probability of disease freedom $PostPFree$ (see [section 8.2](#calc_postpfree_eval)) and to compare it to the targeted probability of freedom (see [section 8.3](#comp_postpfree_eval)).

Finally, the results will be exported (see [chapter 9](#export_eval)).
